Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-14 paper.tex
Save a line
Move deletion before examples
Edit first example

\paragraph{Partial Match State.}
We associate to each goal in Aesop's search tree a partial match state that reflects the goal's context.
Let $\States$ be the set of partial match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
The partial match state's interface then consists of two functions:
\begin{align*}
  \addHyp &: \States → \Hyps → \States × \powerset(\Rules × \CMatches) \\
  \delHyp &: \States → \Hyps → \States
\end{align*}
The $\addHyp$ function adds a hypothesis to the partial match state.
Hence, if $s$ is the partial match state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', M)$, then $s'$ is the partial match state corresponding to $Γ,\, h : T$.
Similarly, if $s$ is the partial match state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the partial match state corresponding to $Γ$.
The second output of $\addHyp$, $M$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
These are the matches that were completed by adding $h : T$.

@@

\paragraph{Example 1.}
\xcom{Technically, it is useless to save level 1 hyps in $\eta_x$. Should we remove them from the examples?.} \jcom{Yes, since the previous section now contains this optimisation.}
To demonstrate the typical behavior of the algorithm we consider the following example which has \textbf{Lemma 1} as a rule and for which we add, in order, the hypotheses $h_1 : 301 \leq 302$ and $h_2 : 300 \leq 301$.
Note that this order is choosed only for demonstration purposes.

At the beginning, we feed $h_1$ to Algorithm \ref{alg:addhyp}.
Furthmore, for the rule \textit{le\_trans}, we notice that the only shared variable is ${y}$.
Now the hypothesis $h_1 : 301 \leq 302$ unifies both with slot 1: $x \leq y$ and slot 2: $y \leq z$.
Starting with slot 1, this updates the map $\eta_y(1, y \mapsto 302)$ with $\{h_1\}$.
An update in slot 1 also marks the begining of a new partial match and we update de match map $\mu_y$ acordingly.
Note that after the creation of a match of level 1, we look at the hypotheses in slot 2 and check if they can extend the match to level 2 (Algorithm \ref{alg:addmatch}). However, the instantiation of the current hypotheses in slot 2 are inconsistent with $y \mapsto 302$, so nothing can be extended. Indeed,
\[ %\textstyle - \nolimits 
  \bigcap_{x\in s(lvl(m_0) + 1)} \eta_x \left(i + 1, \sigma_{m_0}(x)\right) = \eta_y(2, 302) = \emptyset.
\]
Finally, as the hypothesis also matches slot 2, we update the map $\eta_y(2, y \mapsto 301)$ with $\{h_1\}$.
At this point the intersection
\[
  \bigcap_{x\in s(2)} \eta_x \left(1, \sigma_{h}(x)\right) = \mu_y(1, 301) = \emptyset
\]
and at this point the algorithm is done with $h_1$ and stops.
After these updates, the maps have the following state :

\begin{center}
\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
  \node[xshift = 0] {$\eta_y$}
    child[xshift = -20] {node {1}
      child {node {$302$}
        child {node {$\{h_1\}$}}}
    %  child {node {$301$}
    %    child {node {$\{\}$}}}
        }
    child[xshift = 20] {node {2}
      child {node {$301$}
        child {node {$\{h_1\}$}}}
    %  child {node {$300$}
    %    child {node {$\{\}$}}}
    };
  \node[xshift=160] {$\mu_y$}
    child {node[xshift=-20] {1}
      child {node {$302$}
        child {node {$\{[h_1]\}$}}}
    %  child {node {$301$}
    %    child {node {$\{\}$}}}
    };
    %child {node[xshift=20] {2}
    %  child {node {$301$}
    %    child {node[yshift=-2em] {$\{\}$}}}};
\end{tikzpicture}
\end{center}

The next step is to add $h_2 : 300 \leq 301$ to the maps.
Again, it unifies with both slots 1 and 2; we start by updating the map $\eta_y(1, y \mapsto 301)$ with $\{h_2\}$.
Again, an update in slot 1 causes the creation of a level 1 match.
We then check whether there are suitable hypotheses in slot 2 with
with a compatible instantiation of $y$; indeed we find $\eta_y(2, y \mapsto 301) = \{h_1\}$.
This update the match map with the level 2 match $[h_2,h_1]$.
This is a complete match.
The new hypothesis $h_3 : 300 \leq 302$ is added to the context.

The hypothesis $h_2$, also matches with slot 2.
This causes the update $\eta_y(2, y \mapsto 300)$ with $\{h_2\}$. 
At this point, we need to check for level 1 that would be compatible, there are none and the work with this hypothesis is done.

\begin{center}
    \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
      \node[xshift = 0] {$\eta_y$}
        child[xshift = -20] {node {1}
          child {node {$302$}
            child {node {$\{h_1\}$}}}
          child {node {$301$}
            child {node {$\{h_2\}$}}}
            }
        child[xshift = 20] {node {2}
          child {node {$301$}
            child {node {$\{h_1\}$}}}
          child {node {$300$}
            child {node {$\{h_2\}$}}}
        };
      \node[xshift=160] {$\mu_y$}
        child {node[xshift=-20] {1}
          child[xshift=-10] {node {$302$}
            child {node {$\{[h_1]\}$}}}
          child[xshift=10] {node {$301$}
            child {node {$\{[h_2]\}$}}}
        }
        child {node[xshift=20] {2}
          child {node {$301$}
            child {node {$\{[h_2,h_1]\}$}}}};
    \end{tikzpicture}
    \end{center}

>>>>

\paragraph{Partial Match State.}
We associate to each goal in Aesop's search tree a partial match state that reflects the goal's context.
Let $\States$ be the set of partial match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
The partial match state's interface then consists of two functions
\begin{equation*}
  \addHyp : \States → \Hyps → \States × \powerset(\Rules × \CMatches) \qquad \delHyp : \States → \Hyps → \States
\end{equation*}
The $\addHyp$ function adds a hypothesis to the partial match state.
Hence, if $s$ is the partial match state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', M)$, then $s'$ is the partial match state corresponding to $Γ,\, h : T$.
Similarly, if $s$ is the partial match state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the partial match state corresponding to $Γ$.
The second output of $\addHyp$, $M$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
These are the matches that were completed by adding $h : T$.

@@

\paragraph{Example 1.}
As a first example, consider a rule state for the transitivity rule $r : \All{x~y~z}{x ≤ y → y ≤ z → x ≤ z}$ to which we add, in order, the hypotheses
$h₁ : 301 ≤ 302$ and $h₂ : 300 ≤ 301$.
The rule has two slots of types $x ≤ y$ and $y ≤ z$ which share the variable $y$.

The hypothesis $h₁ : 301 ≤ 302$ unifies with both slots, so we run Algorithm~\ref{alg:addhyp} twice.
For slot 1, unification yields the substitution $\{x ↦ 301, y ↦ 302\}$.
Since this is the first slot, we start a new partial match $\{1 ↦ h₁\}$, which Algorithm~\ref{alg:addmatch} inserts into the set $M(y, 1, 302)$.
This match cannot be extended since the hypothesis index $H$ does not contain a hypothesis for slot 2.

For slot 2, unification yields the substitution $σ ≔ \{y ↦ 301, z ↦ 302\}$, so we insert $h₁$ into $H(2, y, 301)$.
No matches at level 1 can be extended by $h₁$ since
\[
  \bigcap_{x \in \vars(2)} M(x, 1, σ(x)) = M(y, 1, 301) = \emptyset
\]
At this point, $H$ and $M$ are in the following state.
For brevity, we write matches $\{1 ↦ h₁, \dots, n ↦ hₙ\}$ as $[h₁, \dots, hₙ]$.
We also write $H_{y}$ instead of $H(y, \text{--}, \text{--})$, and similar for $M$.
\begin{center}
\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
  \node[xshift = 0] {$H_{y}$}
    child {node {2}
      child {node {$301$}
        child {node {$\{h₁\}$}}}
    %  child {node {$300$}
    %    child {node {$\{\}$}}}
    };
  \node[xshift=160] {$M_y$}
    child {node {1}
      child {node {$302$}
        child {node {$\{[h₁]\}$}}}
    %  child {node {$301$}
    %    child {node {$\{\}$}}}
    };
    %child {node[xshift=20] {2}
    %  child {node {$301$}
    %    child {node[yshift=-2em] {$\{\}$}}}};
\end{tikzpicture}
\end{center}

Now we add $h₂ : 300 ≤ 301$, which again unifies with both slots.
For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $\{1 ↦ h₂\}$ to $M(y, 1, 301)$.
We then check for hypotheses in slot 2 that also instantiate $y$ with $301$, and indeed we find $M(2, y, 301) = \{h₁\}$.
Hence we return the complete match $\{1 ↦ h₂, 2 ↦ h₁\}$, which induces a new hypothesis $h₃ : 300 ≤ 302$ to be added later.

For slot 2, we add $h₂$ to $H(y, 2, 300)$.
There are no compatible matches at level 1, so we are done with $h₂$, yielding the following final rule state.
\begin{center}
    \begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
      \node[xshift = 0] {$H_y$}
        child {node {2}
          child {node {$301$}
            child {node {$\{h₁\}$}}}
          child {node {$300$}
            child {node {$\{h₂\}$}}}
        };
      \node[xshift=160] {$M_y$}
        child {node {1}
          child[xshift=-20] {node {$302$}
            child {node {$\{[h₁]\}$}}}
          child[xshift=20] {node {$301$}
            child {node {$\{[h₂]\}$}}}
        };
    \end{tikzpicture}
    \end{center}
